---
title: "Phylogenetic trees -- can we bootstrap?"
author: "Ralph Møller Trane"
date: "10/21/2018"
output: 
  beamer_presentation:
    theme: "Copenhagen"
    colortheme: "beaver"
    slide_level: 2
#classoption: "handout"
---

```{r, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.height = 4, fig.width = 8)
library(ggtree)
library(ggplot2)
library(Biostrings)
library(dplyr)
library(kableExtra)
library(tidyr)
library(stringr)
```

## Phylo-what trees...?

Given a number of aligned DNA sequences, determine how they are related.

Our toy example: cats and dogs!

```{r results='asis'}

cats_dogs_data <- readr::read_delim(file = "../cats-dogs-new/mrbayes_out/cats-dogs.nex", skip = 7, col_names = FALSE, delim = ';', n_max = 12) 

cats_dogs_data %>% 
  #filter(row_number() %in% c(1,2,5,7,8,12)) %>% 
  separate(col = X1, sep = ' ', into = c('Taxa', 'Aligned DNA sequence'), extra = 'merge') %>% 
  mutate(`Aligned DNA sequence` = str_trim(`Aligned DNA sequence`) %>% 
           str_sub(end = 15) %>% 
           paste0("...")) %>% 
  knitr::kable() %>% 
  kableExtra::kable_styling(font_size = 8)
```


## But why?

>- To conserve phylogenetic diversity of plant communities^[https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3440956/], which in turn could lead to feature diversity
>- A certain natural product might have desired qualities, but come with unwanted side effects. 
>- Phylogenetic trees can be related to transmission trees of infectious disease outbreaks^[https://www.ncbi.nlm.nih.gov/pubmed/24037268]
>- Looking for new species
>- Forensics; for example finding evidence that victims were infected by same strain of HIV^[https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4072429/]

## But how?

>1) Based on the observed data, fit a model. 
>2) Sample from the model. 
>3) Results often represented using a single tree annotated with posterior probabilities of clades. 

## Example - cats and dogs

```{r}
tree <- read.tree(text = "((((((gray_wolf,dog),coyote),dhole),racoon_dog),red_fox),(((tiger,(snow_leopard,leopard)),clouded_leopard),(cheetah,cat)));")

probs <- tibble(edge = c(0,tree$edge[,2]),
       prob = case_when(edge == 9 ~ 1,
                        edge == 10 ~ 0.925952,
                        edge == 12 ~ 0.76440,
                        edge == 23 ~ 0.64942,
                        edge == 11 ~ 0.80915,
                        edge == 20 ~ 1,
                        edge == 21 ~ 0.92857,
                        edge == 7 ~ 0.99952,
                        edge == 22 ~ 0.98995,
                        edge == 8 ~ 1,
                        TRUE ~ NA_real_) %>% 
         sprintf(fmt = "%.4f") %>% 
         (function(x = .){if_else(x == "NA", '', x)}),
       clade_number = case_when(edge == 9 ~ "C_2",
                                edge == 10 ~ "C_3",
                                edge == 12 ~ "C_6",
                                edge == 23 ~ "C_4",
                                edge == 11 ~ "C_5",
                                edge == 20 ~ "C_7",
                                edge == 21 ~ "C_8",
                                edge == 7 ~ "C_9",
                                edge == 22 ~ "C_10",
                                edge == 8 ~ "C_11",
                                edge == 19 ~ "C_1", 
                                TRUE ~ NA_character_)) 

tree <- groupClade(tree, .node = c(23,19))


plotted_tree <- ggtree(tree, branch.length = 'none',
                       aes(color = group),
                       right = F) + 
  geom_tiplab(geom = 'text') + 
  geom_nodelab(aes(label = probs$prob), hjust = 1.2, vjust = -0.25, color = 'black') +
  geom_nodelab(aes(label = probs$clade_number), hjust = -0.1, vjust = -0.25, color = 'black') + 
  scale_color_manual(values = c('red', 'blue')) +
  xlim_tree(c(NA, 8))

plotted_tree
```

---

## So... what's the problem?

>- The sample space is HUGE! 654,729,075 possible trees with only 12 taxa
>- When doing inference, we consider the probability of a tree/subclades. 
>    - Estimated using simple relative frequencies (SRF)
>    - Unsampled tree $\implies$ estimated probability is 0. Reasonable?
>- Using Conditional Clade Probabilities (CCD) we can estimate the probability of a tree based on subclades.


## Conditional Clade Probabilities? 

>- Given a split, further evolution in one subclade independent on the other
>- Estimate probability of a tree using bayes and conditional independence

---

```{r}
plotted_tree
```

## Conditional Clade Probabilities? 

- Given a split, further evolution in one subclade independent on the other
- Estimate probability of a tree using bayes and conditional independence

>- SRF based vs CCD based probabilities
>    - similar when we want them to be
>    - CCDs non-zero for unsampled trees


## Where are we going with this? 

>- Don't need to directly estimate tree probabilities, can "simply" estimate CCD
>- Maybe we can do this using bootstrap samples?
>    - pro: much faster!
>    - con: not as accurate
>- Question: if it is less accurate, why bother?
>    - Answer: actually works...
>    - ... sometimes

## Goal:

>- Investigate when it works/doesn't work

## Done so far: 

>- Research. 
>- Learned about phylogenetic trees
>- Tools: MrBayes, Bistro, ccdprobs
>- implemented method to read in a tree in Julia 

## Plan of attack: 

>- simulate trees in Julia
>- use MrBayes/bistro and ccdprobs to get CCDs for bootstrap and MCMC samples
>- compare
>- ~~hopefully~~ find a pattern


